#KIDNEY RENAL CLEAR-CELL-CARCINOMA EXPRESSION ANALYSE

##### Andreu Bofill Pumarla (andreu.bofill01@estudiant.upf.edu)
##### Sergio Castillo Lara (sergio.castillo01@estudiant.upf.edu)
##### Adrià Pérez Culubret (adria.perez06@estudiant.upf.edu)

## Table of Contents

- [LOADING DATA](#loading-data)
	- [Explore phenotypic data](#exploring-phenotypic-data)
	- [Explore feature data](#exploring-feature-data)
	- [DGE object](#dge-object)
- [QUALITY & NORMALIZATION](#quality-&-normalization)
	- [Library size filtering](#library-size-filtering)
	- [Genes Filtering: Distribution of expression levels](#genes-filtering-distribution-of-expression-levels)
		- [Expression levels comparison between all sampes](#expression-levels-comparison-between-all-samples)
    - [Expression levels comparison between normal and tumor](#expression-levels-comparison-between-normal-and-tumor)
  - [Distribution of expression levels among genes](#distribution-of-expression-levels-among-genes)
	- [MA-plots](#ma-plots)
	- [Batch identification](#batch-identification)
    - [GENDER batch effect](#gender-batch-effect)
    - [TSS batch effect](#tss-batch-effect)
    - [PLATE batch effect](#plate-batch-effect)
    - [PORTION-ANALYTE batch effect](#portion-analyte-batch-effect)
- [DIFFERENTIAL EXPRESSION ANALYSIS](#differential-expression-analysis)
  - [Mean Variance relationship](#MEAN-VARIANCE RELATIONSHIP)
  - [Surrogate variable analysis](#surrogate-variable-analysis)
- [CONCLUSIONS](#conclusions)

##LOADING DATA

```{r}
library(SummarizedExperiment)
library(limma)
library(edgeR)
library(sva)
library(ggplot2)
library(geneplotter)
library(org.Hs.eg.db)
library(GOstats)
library(biomaRt)
library(KEGG.db)
```

We load the raw RNA-seq data counts set of Kidney renal clear-cell-carcinoma.

```{r}
se <- readRDS(file.path("~/Documents/Github/IEO-Kidney-expression-analysis/seKIRC.rds"))

se
```

###Exploring phenotypic data

We take a look to the column data, which include the phenotipic information of all samples of the sumarized experiment object.

```{r}
dim(colData(se))
colData(se)[1:5, 1:5]
```

We also need to observe to the metadata information content of these data.

```{r}
mcols(colData(se), use.names=TRUE)
```

As we can see there are three different columns on the metadata result. The first column is the name of all the variables; the second is the labelDescription, which is the definition of each variable; and the last one corresponding to the Common Data Element Identifier (CDEID).

###Exploring feature data

We look through the data rows, which are the feature elements, to see the genes information of our data set.

```{r}
rowRanges(se)
```

###DGE object

We create an object to hold the dataset to be analysed in a better and more comprehensive way.

```{r}
library(edgeR)
dge <- DGEList(counts=assays(se)$counts, genes=mcols(se))
```

Moreover, we calculate $\log_2$ CPM values of expression and we save them in an assay element.

```{r}
assays(se)$logCPM <- cpm(dge, log=TRUE, prior.count=0.5)
```

##QUALITY & NORMALIZATION

##Library size filtering

At this point, we need to determine if the library sizes of tumor samples and normal samples are similar or not.

```{r}
libsize <- data.frame(libsize = dge$sample$lib.size/1e6, type = se$type)

summary(libsize)


ggplot(libsize) +
  geom_density(aes(x=libsize, fill=type), alpha=0.5) +
  ylab("density\n") + xlab("\nMillions of Reads") +
  theme_bw()
```

This past figure shows the density distribution on sequencing depth (milions of reads).

```{r}
ggplot(libsize) +
  geom_histogram(
    aes(x=libsize, fill=type, y = (..count..)/sum(..count..)), alpha=0.5, binwidth=2
  ) + xlab("\nMillions of Reads") + ylab("% of Samples\n") + theme_bw()
```

In this other figure, we can see now the proportion of samples in each sequencing depth (Milions of Reads).

As we can see, both figures show that the normal and tumor samples are similar in terms of sequencing depth, except for some samples that show extreme values.

Moreover, we can see how both normal and tumor types, follow a normal distribution through the sequencing depth.

At this point, we can filter out those samples with a low sequencing depth (<45 Milions of reads) because they are less reliable. 

```{r}
dge.filt <- dge[, dge$sample$lib.size/1e6 > 45 ]

final.samples <- rownames(dge.filt$samples)

se.filt <-se[,colnames(se) %in% final.samples]
```

```{r}
se.normal <- se.filt[,se.filt$type == "normal"]
se.tumor  <- se.filt[,se.filt$type == "tumor"]


normal.code <- substr(colnames(se.normal), 9, 12)
tumor.code <- substr(colnames(se.tumor), 9, 12)

common.codes <- intersect(normal.code, tumor.code)
length(common.codes)

se.paired <- se.filt[,substr(colnames(se.filt), 9, 12) %in% common.codes]
summary(se.paired$type)
colData(se.paired)$samplecodes <- substr(colnames(se.paired), 9, 12)
```


```{r}
dge.filt <- DGEList(counts=assays(se.paired)$counts, genes=mcols(se.paired))
dge.filt
```

```{r libsizes, echo=FALSE, out.width="600px", fig.cap="Figure S1: Library sizes in increasing order."}
ord <- order(dge.filt$sample$lib.size/1e6)
barplot(dge.filt$sample$lib.size[ord]/1e6, las=1, ylab="Millions of reads",
        xlab="Samples", col=c("blue", "red")[(se.paired$type[ord] == "tumor") + 1])
legend("topleft", c("tumor", "normal"), fill=c("red", "blue"), inset=0.01)
```

###Genes filtering: Distribution of expression levels
```{r}

par(mfrow=c(1,2), mar=c(4,5,1,1))
```

####Expression levels comparison between all samples
In the next part, we take a look on the distribution of expression values per sample in terms of logarithmic CPM units.

```{r}
multidensity(as.list(as.data.frame(assays(se.paired)$logCPM)), xlab = "log2 CPM", legend = NULL, main = "All samples", cex.axis = 1.2, cex.lab = 1.5, las = 1)
```

We cannot observe significant diferences between samples in terms of expression levels.

As we have more than 200 samples, we now display the normal and tumor distribution separately.

####Expression levels comparison between normal and tumor

```{r}
normalCPM <- se.paired[,se.paired$type == "normal"]
tumorCPM <- se.paired[,se.paired$type == "tumor"]
multidensity(as.list(as.data.frame(assays(normalCPM)$logCPM)), xlab = "log2 CPM", legend = NULL, main = "Normal", cex.axis = 1.2, cex.lab = 1.5, las = 1)
multidensity(as.list(as.data.frame(assays(tumorCPM)$logCPM)), xlab = "log2 CPM", legend = NULL, main = "Tumor", cex.axis = 1.2, cex.lab = 1.5, las = 1)
```

####Distribution of expression levels among genes

Now, we want to filter out the genes with a low expression. To do that, we can plot the CPMs in logarithmic scale.

```{r}
genemean <- data.frame(Means=rowMeans(assays(se.paired)$logCPM[,-1]))

ggplot(genemean) +
  geom_bar(aes(x=Means), binwidth = 1, fill="white", color="black") +
  theme_bw() + xlab("\nlog2CPM") + ylab("Count\n") +
  geom_vline(xintercept=mean(genemean$Means), color="red")
```


We can see how there are two peaks of genes. The first one correspond to these genes with a very low expression and thus we decide to remove them. In other to do that, we used the mean as a threshold.

```{r}
avgexp <- rowMeans(assays(se.paired)$logCPM)
mask <- avgexp > 1
```

These are the numbers of genes before filtering:

```{r}
dim(se.paired) # Before filtering SE
dim(dge.filt) # Before filtering DGE
```

These are the numbers of samples genes after filtering:

```{r}
se.paired.genes <- se.paired[mask, ]
dim(se.paired.genes) # After filtering SE

dge.filt <- dge.filt[mask, ]
dim(dge.filt) # After filtering DGE

```

At this point, we can calculate the normaliation factors on the filtered expression data set.

```{r}
dgenorm <- calcNormFactors(dge.filt)
assays(se.paired.genes)$logCPMnorm <- cpm(dgenorm, log=TRUE, prior.count=0.5)
```

###MA-plots

We look at the MA-plots of the normal samples to see if there are systematic biases in gene expression levels in any of the sampes. We expect the slope to be around zero.

```{r maPlotsNormal, fig.height=28, fig.width=10, dpi=100, echo=FALSE, fig.cap="Figure S4: MA-plots of the normal samples."}
par(mfrow=c(9, 4 ), mar=c(4, 4, 3, 1))
setmp <- se.paired.genes[, se.paired.genes$type == "normal"]
dgetmp <- dge.filt[, se.paired.genes$type == "normal"]
for (i in 1:ncol(setmp)) {
  A <- rowMeans(assays(setmp)$logCPMnorm)
  M <- assays(setmp)$logCPMnorm[, i] - A
  samplename <- substr(as.character(setmp$bcr_patient_barcode[i]), 1, 12)
  smoothScatter(A, M, main=samplename, las=1)
  abline(h=0, col="blue", lwd=2)
  lo <- lowess(M ~ A)
  lines(lo$x, lo$y, col="red", lwd=2)
}
```

As we can see, we don't see any samples with major biases in gene expression. Now, we examine the tumor samples:

```{r maPlotsTumor, fig.height=110, fig.width=10, dpi=100, echo=FALSE, fig.cap="Figure S4: MA-plots of the tumor samples."}
par(mfrow=c(36, 4 ), mar=c(2, 2, 1, 1))
setmp <- se.paired.genes[, se.paired.genes$type == "tumor"]
dgetmp <- dgenorm[, se.paired.genes$type == "tumor"]
for (i in 1:ncol(setmp)) {
  A <- rowMeans(assays(setmp)$logCPMnorm)
  M <- assays(setmp)$logCPMnorm[, i] - A
  samplename <- substr(as.character(setmp$bcr_patient_barcode[i]), 1, 12)
  smoothScatter(A, M, main=samplename, las=1)
  abline(h=0, col="blue", lwd=2)
  lo <- lowess(M ~ A)
  lines(lo$x, lo$y, col="red", lwd=2)
}
```

###Batch identification

Now we're going to analyze our data in order to search for batch effect that could interfere with the biological signal. First, we analyze some of the information contained in the barcode, such as tissue source site, center of sequenciation, plate and portion and analyte combinations. We use two approaches to try to identify batch effect, the Hierarchical Clustering and a Multidimensional Scaling plot.

```{r}
tss <- substr(colnames(se.paired.genes), 6, 7)
table(tss)
center <- substr(colnames(se.paired.genes), 27, 28)
table(center)
plate <- substr(colnames(se.paired.genes), 22, 25)
table(plate)
portionanalyte <- substr(colnames(se.paired.genes), 18, 20)
table(portionanalyte)
samplevial <- substr(colnames(se.paired.genes), 14, 16)
table(samplevial)


```

####GENDER batch effect

```{r}
table(data.frame(TYPE=se.paired.genes$type, GENDER=se.paired.genes$gender))
```


```{r sampleClusteringGENDER, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S6: Hierarchical clustering of the samples."}
#logCPM <- cpm(dgenorm, log=TRUE, prior.count=3)
d <- as.dist(1-cor(assays(se.paired.genes)$logCPMnorm, method="spearman"))
sampleClustering <- hclust(d)
batch <- as.integer(factor(se.paired.genes$gender))
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)
names(batch) <- colnames(se.paired.genes)
outcome <- paste(substr(colnames(se.paired.genes), 9, 12), as.character(se.paired.genes$type), sep="-")
names(outcome) <- colnames(se.paired.genes)
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)
plot(sampleDendrogram, main="Hierarchical clustering of samples")
legend("topright", paste("TSS", sort(unique(batch)), levels(factor(se.paired.genes$gender))), fill=sort(unique(batch)))

```
```{r mdsGENDER, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S7: Multidimensional scaling plot of the samples."}
plotMDS(dgenorm, labels=outcome, col=batch)
legend("bottomleft", paste("Gender", sort(unique(batch)), levels(factor(se.paired.genes$gender))),
       fill=sort(unique(batch)), inset=0.05)
```

####TSS batch effect

```{r}
table(data.frame(TYPE=se.paired.genes$type, TSS=tss))
```
```{r sampleClusteringTSS, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S6: Hierarchical clustering of the samples."}
#logCPM <- cpm(dgenorm, log=TRUE, prior.count=3)
d <- as.dist(1-cor(assays(se.paired.genes)$logCPMnorm, method="spearman"))
sampleClustering <- hclust(d)
batch <- as.integer(factor(tss))
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)
names(batch) <- colnames(se.paired.genes)
outcome <- paste(substr(colnames(se.paired.genes), 9, 12), as.character(se.paired.genes$type), sep="-")
names(outcome) <- colnames(se.paired.genes)
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)
plot(sampleDendrogram, main="Hierarchical clustering of samples")
legend("topright", paste("TSS", sort(unique(batch)), levels(factor(tss))), fill=sort(unique(batch)))

```
```{r mdsTSS, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S7: Multidimensional scaling plot of the samples."}
plotMDS(dgenorm, labels=outcome, col=batch)
legend("bottomleft", paste("TSS", sort(unique(batch)), levels(factor(tss))),
       fill=sort(unique(batch)), inset=0.05)
```

####PLATE batch effect

```{r}
table(data.frame(TYPE=se.paired.genes$type, PLATE=plate))
```

```{r sampleClusteringPLATE, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S6: Hierarchical clustering of the samples."}
#logCPM <- cpm(dgenorm, log=TRUE, prior.count=3)
d <- as.dist(1-cor(assays(se.paired.genes)$logCPMnorm, method="spearman"))
sampleClustering <- hclust(d)
batch <- as.integer(factor(plate))
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)
names(batch) <- colnames(se.paired.genes)
outcome <- paste(substr(colnames(se.paired.genes), 9, 12), as.character(se.paired.genes$type), sep="-")
names(outcome) <- colnames(se.paired.genes)
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)
plot(sampleDendrogram, main="Hierarchical clustering of samples")
legend("topright", paste("PLATE", sort(unique(batch)), levels(factor(plate))), fill=sort(unique(batch)))

```
```{r mdsPLATE, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S7: Multidimensional scaling plot of the samples."}
plotMDS(dgenorm, labels=outcome, col=batch)
legend("bottomleft", paste("plate", sort(unique(batch)), levels(factor(plate))),
       fill=sort(unique(batch)), inset=0.05)
```

####PORTION-ANALYTE batch effect

```{r}
table(data.frame(TYPE=se.paired.genes$type, P.analyte = portionanalyte))
```
```{r sampleClusteringPortionanalyte, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S6: Hierarchical clustering of the samples."}
#logCPM <- cpm(dgenorm, log=TRUE, prior.count=3)
d <- as.dist(1-cor(assays(se.paired.genes)$logCPMnorm, method="spearman"))
sampleClustering <- hclust(d)
batch <- as.integer(factor(portionanalyte))
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)
names(batch) <- colnames(se.paired.genes)
outcome <- paste(substr(colnames(se.paired.genes), 9, 12), as.character(se.paired.genes$type), sep="-")
names(outcome) <- colnames(se.paired.genes)
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)
plot(sampleDendrogram, main="Hierarchical clustering of samples")
legend("topright", paste("Portionanalyte", sort(unique(batch)), levels(factor(portionanalyte))), fill=sort(unique(batch)))

```
```{r mdsPortionanalyte, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S7: Multidimensional scaling plot of the samples."}
plotMDS(dgenorm, labels=outcome, col=batch)
legend("bottomleft", paste("Portionanalyte", sort(unique(batch)), levels(factor(portionanalyte))),
       fill=sort(unique(batch)), inset=0.05)

```

The Hierarchical clustering and the Multidimensional Scaling plots show a clear differentiation between normal samples and tumor samples. All four plots from the different elements of the barcode have a similar clustering and distribution. There is no clear effect from the batch indicators in the clustering, and it seems to have a balanced design (more clearly in tumor than in normal), although it seems that 9 tumor samples cluster apart from the other samples in the hierarchical clustering, so we should consider taking apart this samples.


## DIFFERENTIAL EXPRESSION ANALYSIS

### MEAN-VARIANCE RELATIONSHIP
```{r}

design <- model.matrix(~type + samplecodes, data = colData(se.paired.genes))
summary(design)
dgenorm
v <- voom(dgenorm, design, plot=FALSE)
FDRcutoff <- 0.01
mod0 <- model.matrix(~samplecodes, colData(se.paired.genes))
sv <- sva(v$E, mod = design, mod0 = mod0)
design.voom <- ""
design.voom <- cbind(design, sv$sv)


fit <- lmFit(v, design.voom)
fit <- eBayes(fit)
res <- decideTests(fit, p.value = FDRcutoff)
summary(res)
volcanoplot(fit, coef = 2, highlight = 7, fit$genes$symbol, main = "Model", las = 1)
colnames(design)

```

```{r}
toptable <- topTable(fit, coef = 2, n=Inf)
par(mfrow=c(1,2), mar=c(4,5,2,2))
hist(toptable$adj.P.Val, xlab="Raw P-values", main = "", las = 1)
qqt(fit$t[,2], df=fit$df.prior + fit$df.residual, main="", pch = ".", cex=3, ylim = c(-50, 50))
abline(0,1, lwd = 2)
```

### OVER vs UNDER
```{r}
summary(toptable)

ggplot(toptable) + geom_histogram(aes(x=logFC), binwidth = 0.01, fill = "#6785d0") + theme_bw()

under_exp <- toptable[toptable$logFC <= -5 ,]
over_exp  <- toptable[toptable$logFC >= 5,] 

under_exp.all <- toptable[toptable$logFC <= 0 ,]
over_exp.all  <- toptable[toptable$logFC >= 0,] 

dim(under_exp)
dim(over_exp)

head(over_exp[order(over_exp$logFC, decreasing = TRUE),])
head(under_exp[order(under_exp$logFC, decreasing = FALSE),])

toptable[toptable$symbol == "PBRM1",]

# PBRM1 is over y es BIEN.
```

### Functional Enrichment 

#### Over represented GOs

```{r}

DEgenes <- rownames(toptable)[toptable$adj.P.Val < 0.01]
length(DEgenes)
geneUniverse <- rownames(se.paired.genes)
params <- new("GOHyperGParams", geneIds=DEgenes, universeGeneIds=geneUniverse,
            annotation="org.Hs.eg.db", ontology="BP",
            pvalueCutoff=0.01, testDirection="over")

conditional(params) <- TRUE
hgOver <- hyperGTest(params)
GOresults <- summary(hgOver)
GOresults <- GOresults[GOresults$Size >= 5 & GOresults$Count >= 5,]

GOresults <- GOresults[order(GOresults$OddsRatio, decreasing = TRUE), ]
GOresults <- GOresults[order(GOresults$Pvalue, decreasing = FALSE), ]
```

head(toptable)

#### Under represented GOs

# (...)

#### Over represented KEGGs

```{r}
DEgenes <- rownames(toptable)[toptable$adj.P.Val < 0.01]
length(DEgenes)
geneUniverse <- rownames(se.paired.genes)
params <- new("KEGGHyperGParams", geneIds=DEgenes, universeGeneIds=geneUniverse,
            annotation="org.Hs.eg.db", pvalueCutoff=0.01, testDirection="over")

hgOver <- hyperGTest(params)

hgOver
KEGGresults <- summary(hgOver)

# Filtering by counts and size
KEGGresults <- KEGGresults[KEGGresults$Size >= 5 & KEGGresults$Count >= 5, ]


KEGGresults <- KEGGresults[order(KEGGresults$OddsRatio, decreasing = TRUE), ]
KEGGresults <- KEGGresults[order(KEGGresults$Pvalue, decreasing = FALSE), ]

head(KEGGresults)
```

When performing the functional enrichment analysis, some fo the gene sets have an OddsRatio value of Infinite. Which is the biological meaning of this?


### GET OMIM ANNOTATIONS
```{r}
over_exp_omim <-  select(
       org.Hs.eg.db,
       keys = rownames(over_exp),
       columns=c("ENTREZID","SYMBOL","GENENAME", "OMIM"),
       keytype="ENTREZID"
)

under_exp_omim <-  select(
       org.Hs.eg.db,
       keys = rownames(under_exp),
       columns=c("ENTREZID","SYMBOL","GENENAME", "OMIM"),
       keytype="ENTREZID"
)


dim(over_exp)
dim(under_exp_omim)

over_exp_omim
under_exp_omim
```

### HIERCHICAL CLUSTERING USING D.E. GENES

Then, we performed a Hierarchical clustering and we plotted a Multidimensional scale plot using only the 77 D.E. genes. 

```{r}
head(colnames(colData(se.paired.genes)))

# Get a mask with DE genes to filter SE object
DE.genes <-  rownames(se.paired.genes) %in% rownames(over_exp) | 
             rownames(se.paired.genes) %in% rownames(under_exp)

# Filter SE object using DE.genes
se.DE  <- se.paired.genes[DE.genes,]
dgenorm.DE <- dgenorm[DE.genes,]


```


```{r sampleClusteringDE, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S6: Hierarchical clustering of the samples."}
#logCPM <- cpm(dgenorm, log=TRUE, prior.count=3)
d <- as.dist(1-cor(assays(se.DE)$logCPMnorm, method="spearman"))
sampleClustering <- hclust(d)
batch <- as.integer(factor(se.DE$type))
sampleDendrogram <- as.dendrogram(sampleClustering, hang=0.1)
names(batch) <- colnames(se.DE)
outcome <- paste(substr(colnames(se.DE), 9, 12), as.character(se.DE$type), sep="-")
names(outcome) <- colnames(se.DE)
sampleDendrogram <- dendrapply(sampleDendrogram,
                               function(x, batch, labels) {
                                 if (is.leaf(x)) {
                                   attr(x, "nodePar") <- list(lab.col=as.vector(batch[attr(x, "label")]))
                                   attr(x, "label") <- as.vector(labels[attr(x, "label")])
                                 }
                                 x
                               }, batch, outcome)
plot(sampleDendrogram, main="Hierarchical clustering of samples")
legend("topright", paste("TYPE", sort(unique(batch)), levels(factor(se.DE$type))), fill=sort(unique(batch)))

```

```{r mdsDE, fig.height=7, fig.width=14, dpi=100, echo=TRUE, fig.cap="Figure S7: Multidimensional scaling plot of the samples."}
plotMDS(dgenorm.DE, labels=outcome, col=batch)
legend("bottomleft", paste("Type", sort(unique(batch)), levels(factor(se.paired.genes$gender))),
       fill=sort(unique(batch)), inset=0.05)
```

##Session information

```{r sessionInfo}
sessionInfo()
```
