
```{r setup, cache=FALSE, echo=FALSE, results='asis'}

```


```
library(knitr)     ## required for "knitting" from Rmd to md
library(markdown)  ## required for processing from md to HTML
knit2html("~/Dropbox/Home/Documentos/MASTER/IEO/KIDNEY/project.Rmd")  ## process Rmd to HTML
browseURL("projectTemplate.html") ## open the resulting HTML file from R
```

## Loading DATA
```{r}
library(SummarizedExperiment)
se <- readRDS(file.path("~/GitHub/KIDNEY/seKIRC.rds"))
se
```

## Explore phenotypic data

```{r}
dim(colData(se))
colData(se)[1:5, 1:5]
mcols(colData(se), use.names=TRUE)
```

## Explore feature data
```{r}
rowRanges(se)
```


## DGE object
```{r}
library(edgeR)

dge <- DGEList(counts=assays(se)$counts, genes=mcols(se))
assays(se)$logCPM <- cpm(dge, log=TRUE, prior.count=0.5)
```


# QUALITY & NORMALIZATION
```{r}
dgenorm <- calcNormFactors(dge)
assays(se)$logCPMnorm <- cpm(dgenorm, log=TRUE, prior.count=0.5)
```

## Remove paired samples

EXPLICACION

```{r}
normal.samples <- se[,se$type == "normal"]
tumor.samples  <- se[,se$type == "tumor"]

dge.normal <- DGEList(counts=assays(normal.samples)$counts, genes=mcols(normal.samples))
dge.tumor  <- DGEList(counts=assays(tumor.samples)$counts, genes=mcols(tumor.samples))

normal.samples <- rownames(dge.normal$samples)
tumor.samples  <- rownames(dge.tumor$samples)

common.samples <- intersect(normal.samples,tumor.samples)
common.samples

dgefilt <- dge[, !rownames(dge$samples) %in% common.samples]
```


## Library size

```{r}
library(ggplot2)

libsize <- data.frame(libsize = dge$sample$lib.size/1e6, type = se$type)

summary(libsize)


ggplot(libsize) + 
  geom_density(aes(x=libsize, fill=type), alpha=0.5) +
  ylab("density\n") + xlab("\nMillions of Reads") +
  theme_bw()
  
ggplot(libsize) + 
  geom_histogram(
    aes(x=libsize, fill=type, y = (..count..)/sum(..count..)), alpha=0.5, binwidth=3
  ) + xlab("\nMillions of Reads") + ylab("% of Samples\n") + theme_bw()

dgefilt <- dgefilt[, dgefilt$sample$lib.size/1e6 > 50 & dgefilt$sample$lib.size/1e6 < 70]
```



```{r}
final.samples <- rownames(dgefilt$samples)

se.filt <-se[,colnames(se) %in% final.samples]
summary(se.filt$samples)
```


#### After filtering
```{r}
ggplot(libsize) + 
  geom_histogram(
    aes(x=libsize, fill=type, y = (..count..)/sum(..count..)), alpha=0.5, binwidth=3
  ) + xlab("\nMillions of Reads") + ylab("% of Samples\n") + theme_bw() + xlim(c(25, 71))

ggplot(libsize) + 
  geom_density(aes(x=libsize, fill=type), alpha=0.5) +
  ylab("density\n") + xlab("\nMillions of Reads") +
  theme_bw() + xlim(c(25, 71))


### NO TOCAR; NEGRO
```{r libsizes, echo=FALSE, out.width="600px", fig.cap="Figure S1: Library sizes in increasing order."}
ord <- order(dge$sample$lib.size/1e6)
barplot(dge$sample$lib.size[ord]/1e6, las=1, ylab="Millions of reads",
        xlab="Samples", col=c("yellow", "red")[(se$type[ord] == "tumor") + 1])
legend("topleft", c("tumor", "normal"), fill=c("red", "yellow"), inset=0.01)
```


### Distribution of expression levels among samples
```{r}
library(geneplotter)
par(mfrow=c(1,2), mar=c(4,5,1,1))
```
### Distribution of expression levels (All samples)
```{r}
multidensity(as.list(as.data.frame(assays(se.filt)$logCPMnorm)), xlab = "log2 CPM", legend = NULL, main = "All samples", cex.axis = 1.2, cex.lab = 1.5, las = 1)
```

## Expression levels comparison between normal and tumor
```{r}
normalCPMnorm <- se.filt[,se.filt$type == "normal"]
tumorCPMnorm <- se.filt[,se.filt$type == "tumor"]
multidensity(as.list(as.data.frame(assays(normalCPMnorm)$logCPMnorm)), xlab = "log2 CPM", legend = NULL, main = "Normal", cex.axis = 1.2, cex.lab = 1.5, las = 1)
multidensity(as.list(as.data.frame(assays(tumorCPMnorm)$logCPMnorm)), xlab = "log2 CPM", legend = NULL, main = "Tumor", cex.axis = 1.2, cex.lab = 1.5, las = 1)
```



